---
title: "pkgstats"
output:
  md_document:
    variant: gfm

  rmarkdown::html_vignette:
    self_contained: no
---

<!-- badges: start -->
[![R build status](https://github.com/ropensci-review-tools/pkgstats/workflows/R-CMD-check/badge.svg)](https://github.com/ropensci-review-tools/pkgstats/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/ropensci-review-tools/pkgstats/branch/main/graph/badge.svg)](https://app.codecov.io/gh/ropensci-review-tools/pkgstats)
[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

```{r, include = FALSE, echo = FALSE}
if (identical (Sys.getenv ("IN_PKGDOWN"), "true")) {
    pkgstats::ctags_install ()
}
```


# pkgstats

Extract summary statistics of R package structure and functionality. Not all
statistics of course, but a good go at balancing insightful statistics while
ensuring computational feasibility. `pkgstats` is a *static* code analysis
tool, so is generally very fast (a few seconds at most for very large
packages).

## What statistics?

Statistics are derived from these primary sources:

1. Numbers of lines of code, documentation, and white space (both between and
   within lines) in each directory and language
2. Summaries of package `DESCRIPTION` file and related package meta-statistics
3. Summaries of all objects created via package code across multiple languages
   and all directories containing source code (`./R`, `./src`, and
   `./inst/include`).
4. A function call network derived from function definitions obtained from
   [the code tagging library, `ctags`](https://ctags.io), and references
   ("calls") to those obtained from [another tagging library,
   `gtags`](https://www.gnu.org/software/global/). This network roughly
   connects every object making a call (as `from`) with every object being
   called (`to`).
5. An additional function call network connecting calls within R functions to
   all functions from other R packages.

The [primary function,
`pkgstats()`](https://docs.ropensci.org/pkgstats/reference/pkgstats.html),
returns a list of these various components, including full `data.frame` objects
for the final three components described above. The statistical properties of
this list can be aggregated by the [`pkgstats_summary()`
function](https://docs.ropensci.org/pkgstats/reference/pkgstats_summary.html),
which returns a `data.frame` with a single row of summary statistics. This
function is demonstrated below, including full details of all statistics
extracted.


## Installation

The easiest way to install this package is [via the associated
`r-universe`](https://ropensci-review-tools.r-universe.dev/ui#builds). As
shown there, simply enable the universe with

```{r options, eval = FALSE}
options (repos = c (
    ropenscireviewtools = "https://ropensci-review-tools.r-universe.dev",
    CRAN = "https://cloud.r-project.org"
))
```

And then install the usual way with,

```{r install, eval = FALSE}
install.packages ("pkgstats")
```

Alternatively, the package can be installed by running one of the following
lines:



```{r remotes, eval = FALSE}
remotes::install_github ("ropensci-review-tools/pkgstats")
pak::pkg_install ("ropensci-review-tools/pkgstats")
```

The package can then loaded for use with:

```{r library, eval = TRUE}
library (pkgstats)
```

### Installation on Linux systems

This package requires the [system libraries
`ctags-universal`](https://ctags.io) and [GNU
`global`](https://www.gnu.org/software/global/), both of which are
automatically installed along with the package on both Windows and MacOS
systems. Most Linux distributions do not include a sufficiently up-to-date
version of [`ctags-universal`](https://ctags.io), and so it must be compiled
from source. This can be done by running a single function, `ctags_install()`,
which will install both [`ctags-universal`](https://ctags.io) and [GNU
`global`](https://www.gnu.org/software/global/).

The `pkgstats` package includes a function to ensure your local installations
of `universal-ctags` and `global` work correctly. Please ensure you see the
following prior to proceeding:

```{r ctags-check, eval = TRUE}
ctags_test ()
```

Note that GNU `global` can be linked at installation to the Universal Ctags
plug-in parser to expand the [default 5 languages to
30](https://www.gnu.org/software/global/). This makes no difference to
`pkgstats` results, as `gtags` output is only used to trace function call
networks, which is only possible for compiled languages able to dynamically
share pointers to the same objects. This is possible with the default parser
regardless. The wealth of extra information obtained from linking `global` to
the Universal Ctags parser is ultimately discarded anyway, yet parsing may take
considerably longer. If this is the case, "default" behaviour may be recovered
by first running the following command:

```{r unsetenv, eval = FALSE}
Sys.unsetenv (c ("GTAGSCONF", "GTAGSLABEL"))
```

See [information on how to install the
plugin](https://cvs.savannah.gnu.org/viewvc/global/global/plugin-factory/PLUGIN_HOWTO.pygments?revision=1.6&view=markup)
for more details.


## Demonstration

The following code demonstrates the output of the main function, `pkgstats`,
using an internally bundled `.tar.gz` "tarball" of this package. The
`system.time` call demonstrates that the static code analyses of `pkgstats` are
generally very fast.

```{r demo}
tarball <- system.file ("extdata", "pkgstats_9.9.tar.gz", package = "pkgstats")
system.time (
    p <- pkgstats (tarball)
)
names (p)
```

The result is a list of various data extracted from the code. All except for
`objects` and `network` represent summary data:

```{r}
p [!names (p) %in% c ("objects", "network", "external_calls")]
```

These results demonstrate that many fields use `NA` to denote values of zero.
The first item, `loc`, contains the following Lines-Of-Code and related
statistics, separated into distinct combinations of computer language and
directory:

1. `nfiles` = Numbers of files in each directory and language.
2. `nlines` = Total numbers of lines in all files.
2. `nlines` = Total numbers of lines of code.
3. `ndoc` = Total numbers of documentation or comment lines.
4. `nempty` = Total numbers of empty of blank lines.
5. `nspaces` = Total numbers of white spaces in all code lines, excluding
   leading indentation spaces.
6. `nchars` = Total numbers of non-white-space characters in all code lines.
7. `nexpr` = Median numbers of nested expressions in all lines which have any
   expressions (see below).
8. `ntabs` = Number of lines of code with initial tab indentation.
8. `indentation` = Number of spaces by which code is indented (with `-1`
   denoting tab-indentation).

Numbers of nested expressions are counted as numbers of brackets or braces of
any type nested on a single line. The following line has one nested bracket:

```{r nested1, eval = FALSE}
x <- myfn ()
```

while the following has four:

```{r nested4, eval = FALSE}
x <- function () { return (myfn ()) }
```

Code with fewer nested expressions per line is generally easier to read, and
this metric is provided as one indication of the general readability of
code. A second relative indication may be extracted by converting numbers of
spaces and characters to a measure of relative numbers of white spaces, noting
that the `nchars` value quantifies total characters including white spaces.

```{r rel-space}
index <- which (p$loc$dir %in% c ("R", "src")) # consider source code only
sum (p$loc$nspaces [index]) / sum (p$loc$nchars [index])
```

Finally, the `ntabs` statistic can be used to identify whether code uses tab
characters as indentation, otherwise the `indentation` statistics indicate
median numbers of white spaces by which code is indented. The `objects`,
`network`, and `external_calls` items returned by the [`pkgstats()`
function](https://docs.ropensci.org/pkgstats/reference/pkgstats.html) are
described further below.


### Overview of statistics and the `pkgstats_summary()` function

A summary of the `pkgstats` data can be obtained by submitting the object
returned from `pkgstats()` to the [`pkgstats_summary()`
function](https://docs.ropensci.org/pkgstats/reference/pkgstats_summary.html):

```{r summary, echo = TRUE}
s <- pkgstats_summary (p)
```

This function reduces the result of the [`pkgstats()`
function](https://docs.ropensci.org/pkgstats/reference/pkgstats_summary.html)
to a single line with `r ncol (s)` entries, represented as a `data.frame` with
one row and that number of columns. This format is intended to enable summary
statistics from multiple packages to be aggregated by simply binding rows
together. While `r ncol (s)` statistics might seem like a lot, the
[`pkgstats_summary()`
function](https://docs.ropensci.org/pkgstats/reference/pkgstats_summary.html)
aims to return as many usable raw statistics as possible in order to flexibly
allow higher-level statistics to be derived through combination and
aggregation. These `r ncol (s)` statistics can be roughly grouped into the
following categories (not shown in the order in which they actually appear),
with variable names in parentheses after each description. Some statistics are
summarised as comma-delimited character strings, such as translations into
human languages, or other packages listed under "depends", "imports", or
"suggests". This enables subsequent analyses of their contents, for example of
actual translated languages, or both aggregate numbers and individual details
of all package dependencies, as demonstrated immediately below.

**Package Summaries**

- name (`package`)
- Package version (`version`)
- Package date, as modification time of `DESCRIPTION` file where not explicitly
  stated (`date`)
- License (`license`)
- Languages, as a single comma-separated character value (`languages`), and
  excluding `R` itself.
- List of translations where package includes translations files, given as list
  of (spoken) language codes (`translations`).

**Information from `DESCRIPTION` file**

- Package URL(s) (`url`)
- URL for BugReports (`bugs`)
- Number of contributors with role of *author* (`desc_n_aut`), *contributor*
  (`desc_n_ctb`), *funder* (`desc_n_fnd`), *reviewer* (`desc_n_rev`), *thesis
  advisor* (`ths`), and *translator* (`trl`, relating to translation between
  computer and not spoken languages).
- Comma-separated character entries for all `depends`, `imports`, `suggests`,
  and `linking_to` packages.

Numbers of entries in each the of the last two kinds of items can be obtained
from by a simple `strsplit` call, like this:

```{r strsplit}
deps <- strsplit (s$suggests, ", ") [[1]]
length (deps)
print (deps)
```

**Numbers of files and associated data**

- Number of vignettes (`num_vignettes`)
- Number of demos (`num_demos`)
- Number of data files (`num_data_files`)
- Total size of all package data (`data_size_total`)
- Median size of package data files (`data_size_median`)
- Numbers of files in main sub-directories (`files_R`, `files_src`,
  `files_inst`, `files_vignettes`, `files_tests`), where numbers are
  recursively counted in all sub-directories, and where `inst` only counts
  files in the `inst/include` sub-directory.

**Statistics on lines of code**

- Total lines of code in each sub-directory (`loc_R`, `loc_src`, `loc_ins`,
  `loc_vignettes`, `loc_tests`).
- Total numbers of blank lines in each sub-directory (`blank_lines_R`,
  `blank_lines_src`, `blank_lines_inst`, `blank_lines_vignette`,
  `blank_lines_tests`).
- Total numbers of comment lines in each sub-directory (`comment_lines_R`,
  `comment_lines_src`, `comment_lines_inst`, `comment_lines_vignettes`,
  `comment_lines_tests`).
- Measures of relative white space in each sub-directory (`rel_space_R`,
  `rel_space_src`, `rel_space_inst`, `rel_space_vignettes`, `rel_space_tests`),
  as well as an overall measure for the `R/`, `src/`, and `inst/` directories
  (`rel_space`).
- The number of spaces used to indent code (`indentation`), with values of -1
  indicating indentation with tab characters.
- The median number of nested expression per line of code, counting only those
  lines which have any expressions (`nexpr`).

**Statistics on individual objects (including functions)**

These statistics all refer to "functions", but actually represent more general
"objects," such as global variables or class definitions (generally from
languages other than R), as detailed below.

- Numbers of functions in R (`n_fns_r`)
- Numbers of exported and non-exported R functions (`n_fns_r_exported`,
  `n_fns_r_not_exported`)
- Number of functions (or objects) in other computer languages (`n_fns_src`),
  including functions in both `src` and `inst/include` directories.
- Number of functions (or objects) per individual file in R and in all other
  (`src`) directories (`n_fns_per_file_r`, `n_fns_per_file_src`).
- Median and mean numbers of parameters per exported R function
  (`npars_exported_mn`, `npars_exported_md`).
- Mean and median lines of code per function in R and other languages,
  including distinction between exported and non-exported R functions
  (`loc_per_fn_r_mn`, `loc_per_fn_r_md`, `loc_per_fn_r_exp_m`,
  `loc_per_fn_r_exp_md`, `loc_per_fn_r_not_exp_mn`, `loc_per_fn_r_not_exp_m`,
  `loc_per_fn_src_mn`, `loc_per_fn_src_md`).
- Equivalent mean and median numbers of documentation lines per function
  (`doclines_per_fn_exp_mn`, `doclines_per_fn_exp_md`,
  `doclines_per_fn_not_exp_m`, `doclines_per_fn_not_exp_md`,
  `docchars_per_par_exp_mn`, `docchars_per_par_exp_m`).

**Network Statistics**

The full structure of the `network` table is described below, with summary
statistics including:

- Number of edges, including distinction between languages (`n_edges`,
  `n_edges_r`, `n_edges_src`).
- Number of distinct clusters in package network (`n_clusters`).
- Mean and median centrality of all network edges, calculated from both
  directed and undirected representations of network (`centrality_dir_mn`,
  `centrality_dir_md`, `centrality_undir_mn`, `centrality_undir_md`).
- Equivalent centrality values excluding edges with centrality of zero
  (`centrality_dir_mn_no0`, `centrality_dir_md_no0`, `centrality_undir_mn_no0`,
  `centrality_undir_md_no`).
- Numbers of terminal edges (`num_terminal_edges_dir`,
  `num_terminal_edges_undir`).
- Summary statistics on node degree (`node_degree_mn`, `node_degree_md`,
  `node_degree_max`)


**External Call Statistics**

The final column in the result of [the `pkgstats_summary()`
function](https://docs.ropensci.org/pkgstats/reference/pkgstats_summary.html)
summarises the `external_calls` object detailing all calls make to external
packages (including to base and recommended packages). This summary is
also represented as a single character string. Each package lists total numbers
of function calls, and total numbers of unique function calls. Data for each
package are separated by a comma, while data within each package are separated
by a colon.

```{r summary-external-calls}
s$external_calls
```

This structure allows numbers of calls to all packages to be readily extracted
with code like the following:

```{r summary-exteranl-calls-transform}
calls <- do.call (
    rbind,
    strsplit (strsplit (s$external_call, ",") [[1]], ":")
)
calls <- data.frame (
    package = calls [, 1],
    n_total = as.integer (calls [, 2]),
    n_unique = as.integer (calls [, 3])
)
print (calls)
```

The two numeric columns respectively show the total number of calls made to
each package, and the total number of unique functions used within those
packages. These results provide detailed information on numbers of calls made
to, and functions used from, other R packages, including base and recommended
packages.

The following sub-sections provide further detail on the `objects`, `network`,
and `external_call` items, which could be used to extract additional statistics
beyond those described here.


### Objects

The `objects` item contains all code objects identified by
the code-tagging library [`ctags`](https://ctags.io). For R, those are
primarily functions, but for other languages may be a variety of entities such
as class or structure definitions, or sub-members thereof. Object tables look
like this:

```{r}
head (p$objects)
```

The `magrittr` package has a total of `r nrow (p$objects)` objects, which the
following lines provide some insight into.

```{r}
table (p$objects$language)
table (p$objects$kind)
table (p$objects$kind [p$objects$language == "R"])
table (p$objects$kind [p$objects$language == "C++"])
```

### Network

The `network` item details all relationships between objects, which generally
reflects one object calling or otherwise depending on another object. Each row
thus represents one edge of a "function call" network, with each entry in the
`from` and `to` columns representing the network vertices or nodes.

```{r}
head (p$network)
nrow (p$network)
```

The network table includes additional statistics on the centrality of each
edge, measured as betweenness centrality assuming edges to be both directed
(`centrality_dir`) and undirected (`centrality_undir`). More central edges
reflect connections between objects that are more central to package
functionality, and vice versa. The distinct components of the network are also
represented by discrete cluster numbers, calculated both for directed and
undirected versions of the network. Each distinct cluster number represents
a distinct group of objects, internally related to other members of the same
cluster, yet independent of all objects with different cluster numbers.

The network can be viewed as an interactive [`vis.js`](https://visjs.org/)
network through passing the result of `pkgstats` -- the variable `p` in the
code above -- to the [`plot_network()`
function](https://docs.ropensci.org/pkgstats/reference/plot_network.html).

### External Calls

The `external_calls` item is structured similar to the `network` object, but
identifies all calls to functions from external packages. However, unlike the
`network` and `object` data, which provide information on objects and
relationships in all computer languages used within a package, the
`external_calls` object maps calls within R code only, in order to provide
insight into the use within a package of of functions from other packages,
including R's base and recommended packages. The object looks like this:

```{r ext-call-head}
head (p$external_calls)
```

These data are converted to a summary form by the [`pkgstats_summary()`
function](https://docs.ropensci.org/pkgstats/reference/pkgstats_summary.html),
which tabulates numbers of external calls and unique functions from each
package. These data are presented as a single character string which can be
easily converted to the corresponding numeric values using code like the
following:

```{r ext-call-summary}
x <- strsplit (s$external_calls, ",") [[1]]
x <- do.call (rbind, strsplit (x, ":"))
x <- data.frame (
    pkg = x [, 1],
    n_total = as.integer (x [, 2]),
    n_unique = as.integer (x [, 3])
)
x$n_total_rel <- round (x$n_total / sum (x$n_total), 3)
x$n_unique_rel <- round (x$n_unique / sum (x$n_unique), 3)
print (x)
```

Those data reveal, for example, that the `magrittr` package makes 
`r x$n_total [x$pkg == "base"]` individual calls to
`r x$n_unique [x$pkg == "base"]` unique functions from the "base" package.



## Code of Conduct

Please note that this package is released with a [Contributor Code of
Conduct](https://ropensci.org/code-of-conduct/). By contributing to this
project, you agree to abide by its terms.
