# pkgstats

<!-- badges: start -->
[![R build status](https://github.com/mpadge/pkgstats/workflows/R-CMD-check/badge.svg)](https://github.com/mpadge/pkgstats/actions?query=workflow%3AR-CMD-check)
[![Project Status: Concept](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
<!-- badges: end -->

Historical statistics of every R package ever. Not all statistics of course,
but a good go at balancing insightful statistics while ensuring computational
feasibility. Requires a local mirror of the full CRAN archive - that is, all
[current
packages](https://cran.r-project.org/web/packages/available_packages_by_name.html),
plus all of their [previous
versions](https://cran.r-project.org/src/contrib/Archive).

## Installation

The package can be installed with:
```{r remotes, eval = FALSE}
remotes::install_github ("mpadge/pkgtest")
```
And then loaded for use with
```{r library, eval = FALSE}
library (pkgstats)
```
```{r load_all, echo = FALSE, message = FALSE}
devtools::load_all (export_all = FALSE)
```



This package requires [`ctags`](https://ctags.io), and includes a function to
ensure your local installation works correctly. Please ensure you see the
following prior to proceeding:

```{r ctags-check}
ctags_test ()
```


## Demonstration

```{r demo}
library (pkgstats)
tarball <- "magrittr_2.0.1.tar.gz"
u <- paste0 ("https://cran.r-project.org/src/contrib/",
             tarball)
f <- file.path (tempdir (), tarball)
download.file (u, f)
system.time (
    p <- pkgstats (f)
    )
print (p)
```
