---
title: "Installation"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
options (repos = c (
    ropenscireviewtools = "https://ropensci-review-tools.r-universe.dev",
    CRAN = "https://cloud.r-project.org"
))
```

The easiest way to install this package is [via the associated
`r-universe`](https://ropensci-review-tools.r-universe.dev/ui#builds). As
shown there, simply enable the universe with

```{r options, eval = FALSE}
options (repos = c (
    ropenscireviewtools = "https://ropensci-review-tools.r-universe.dev",
    CRAN = "https://cloud.r-project.org"
))
```

And then install the usual way with,

```{r install, eval = FALSE}
install.packages ("pkgstats")
```

Alternatively, the package can be installed by running one of the following
lines:



```{r remotes, eval = FALSE}
remotes::install_github ("ropensci-review-tools/pkgstats")
pak::pkg_install ("ropensci-review-tools/pkgstats")
```

The package can then loaded for use with:

```{r library, eval = TRUE}
library (pkgstats)
```

## Installation on Linux systems

This package requires the [system libraries
`ctags-universal`](https://ctags.io) and [GNU
`global`](https://www.gnu.org/software/global/), both of which are
automatically installed along with the package on both Windows and MacOS
systems. Most Linux distributions do not include a sufficiently up-to-date
version of [`ctags-universal`](https://ctags.io), and so it must be compiled
from source. This can be done by running a single function, `ctags_install()`,
which will install both [`ctags-universal`](https://ctags.io) and [GNU
`global`](https://www.gnu.org/software/global/).

The `pkgstats` package includes a function to ensure your local installations
of `universal-ctags` and `global` work correctly. Please ensure you see the
following prior to proceeding:

```{r ctags-check, eval = FALSE}
ctags_test ()
#> [1] TRUE
```

Note that GNU `global` can be linked at installation to the Universal Ctags
plug-in parser to expand the [default 5 languages to
30](https://www.gnu.org/software/global/). This makes no difference to
`pkgstats` results, as `gtags` output is only used to trace function call
networks, which is only possible for compiled languages able to dynamically
share pointers to the same objects. This is possible with the default parser
regardless. The wealth of extra information obtained from linking `global` to
the Universal Ctags parser is ultimately discarded anyway, yet parsing may take
considerably longer. If this is the case, "default" behaviour may be recovered
by first running the following command:

```{r unsetenv, eval = FALSE}
Sys.unsetenv (c ("GTAGSCONF", "GTAGSLABEL"))
```

See [information on how to install the
plugin](https://cvs.savannah.gnu.org/viewvc/global/global/plugin-factory/PLUGIN_HOWTO.pygments?revision=1.6&view=markup)
for more details.
